---
title: "Stats 230 - HW1"
author: "Pieter Derdeyn"
date: "1/14/2022"
repo: "https://github.com/pderdeyn/Stats230pieter"
output: html_document
vignette: >
  %\VignetteIndexEntry{my-vignette}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Multiply Some Matrices

Let's multiply matrices. We use a function, `mat_mult()`, from my package available at https://github.com/pderdeyn/Stats230pieter. `mat_mult` computes the product $A\times B\times x$. This function uses an optional boolean argument, `left`, which determines whether the left product, $A\times B$, or the right product, $B \times x$, is computed first. The default argument is `left=TRUE`. 

```{r matrix_mult}
#install.packages("../",repos=NULL,type="source")
#devtools::install_github('https://github.com/pderdeyn/Stats230pieter')
library(Stats230pieter)
values<-seq(0,100)
A<-matrix(sample(values,100),ncol=10)
B<-matrix(sample(values,100),ncol=10)
x<-c(sample(values,10))
mat_mult(A,B,x)
mat_mult(A,B,x,left=FALSE)
```

## Benchmarking

Let's benchmark multiplying matrices

```{r benchmark}
library(bench)
b<-bench::mark(mat_mult(A,B,x,left=TRUE),
            mat_mult(A,B,x,left=FALSE))
plot(b)
```

Hence, it appears from this experiment that computing the right product, $B\times x$, first leads to faster computation. It is interesting to see that this expression has multimodal distribution, where the left product variation has just one peak. 
